---
title: "PhyloseqMRAManuscript"
author: "Tricia"
date: "`r Sys.Date()`"
output: pdf_document
---

##Load required packages
```{r Load Required Packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(here)
library(phyloseq)
library(vegan)
library(dplyr)
library(plyr)
library(decontam)
library(MicEco) #psvenn
library(BiMiCo) #rmnonbac
library(ggpubr) #statcomparemeans
```

##Load taxa and seqtab files to start here
```{r Load DADA2 files}
load("../RData/taxa.RData")
load("../RData/seqtab.nochim.RData")
```

##import metadata
```{r Import Metadata}
metadata<-read.csv("../metadata.csv", header=TRUE, row.names = 1)
```

##Create phyloseq object
```{r Create Phyloseq Object}
#make sure the seqtab.nochim and taxa objects are loaded
physeq <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(metadata), 
               tax_table(taxa))
physeq
```

##inspect library sizes
```{r Inspect Library Size}
df <- as.data.frame(sample_data(physeq)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(physeq)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=control)) + geom_point()
```

##identify contaminants
```{r Identify Contaminants at Higher Freq in Negative Controls, warning=FALSE}
sample_data(physeq)$is.neg <- sample_data(physeq)$control == "yes"
contamdf.prev <- isContaminant(physeq, method="prevalence", neg="is.neg", threshold=0.5) #identify contaminants in higher frequency in negative controls
table(contamdf.prev$contaminant)
head(which(contamdf.prev$contaminant))
```

##remove control samples
```{r Remove Control Samples}
# Make phyloseq object of presence-absence in negative controls and true samples
physeq.pa <- transform_sample_counts(physeq, function(abund) 1*(abund>0))
physeq.pa.neg <- prune_samples(sample_data(physeq.pa)$control == "yes", physeq.pa)
physeq <- prune_samples(sample_data(physeq.pa)$control == "no", physeq.pa) #this will contain positives
physeq
```

##graph controls
```{r Graph Controls}
df.pa <- data.frame(pa.pos=taxa_sums(physeq), pa.neg=taxa_sums(physeq.pa.neg),
                      contaminant=contamdf.prev$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
```

##remove contaminants
```{r Remove Contaminants}
physeq <- prune_taxa(!contamdf.prev$contaminant, physeq)
physeq
```

##Remove mock community
```{r Remove Mock Community}
physeq <- subset_samples(physeq, mock != "yes") 
physeq
```

##Remove the sequence itself and replace with ASV
```{r Separate Sequence Out}
dna <- Biostrings::DNAStringSet(taxa_names(physeq))
names(dna) <- taxa_names(physeq)
physeq <- merge_phyloseq(physeq, dna)
taxa_names(physeq) <- paste0("ASV", seq(ntaxa(physeq)))
physeq
```

##remove mitochondria and chloroplast matches. 
```{r Remove Mito and Chloro}
physeq <- physeq %>% subset_taxa( Family!= "Mitochondria" | is.na(Family) & Order!="Chloroplast" | is.na(Order) ) 
physeq
```

##remove all non bacterial sequences
```{r Remove Non-Bact}
physeq<-rm_nonbac(physeq)
physeq
```

##save physeq object as a file
```{r Save Phyloseq Object}
##save physeq object as R file
save(physeq, file="../RData/physeq.RData")
```

##load physeq
```{r Load Phyloseq Object}
load("../RData/physeq.RData")
```

##Alpha Diversity based on site with stats
```{r Alpha Diversity Graph, message=FALSE, warning=FALSE}
# Plot richness
p <- plot_richness(physeq, x = "site", measures = c("Observed", "Simpson", "Shannon"))

# Create boxplot without statistical tests
barsite <- p + 
  geom_boxplot(data = p$data, aes(x = site, y = value, color = NULL), alpha = 0.1) + xlab("Site") + ggtitle("Alpha Diversity at Each Site") +
  theme(axis.title = element_text(face = "bold"), 
          axis.title.x = element_text(face="bold"), plot.title = element_text(size = 10, face = "bold", hjust = .5))

print(barsite)
```

##export tiff with 300dpi
```{r Export Alpha Diversity Graph}
ggsave(
  filename="../figures/Figure01AlphaDiv.tiff",
  plot = barsite,
  width = 200,
  height = 200,
  units = c("mm"),
  dpi = 300,
)
```

##Export alpha diveristy
```{r Export Alpha Diversity}
alphadiv<-estimate_richness(physeq, measures=c("Observed", "Shannon", "Simpson"))
write.csv(alphadiv, "../alphasheets/alpha_div.csv")
```

##add site info to alpha_div file and rename alphadiv import for normality testing
```{r Import Alpha Diversity with Site Detail}
meta<-read.csv("../alphasheets/alphadiv.csv")
```

##hist
```{r Histograph of Alpha Div}
par(mfrow = c(1, 3))
hist(meta$Shannon, main="Shannon diversity", xlab="", breaks=10)
hist(meta$Simpson, main="Simpson diversity", xlab="", breaks=10)
hist(meta$Observed, main="Observed ASVs", xlab="", breaks=10)
```

##Test for normality (Shapiro)
```{r Test for Normality}
shapiro.test(meta$Shannon)
shapiro.test(meta$Simpson)
shapiro.test(meta$Observed) #normal
```

##Two factor tests for paired samples
```{r Two-factor Tests}
##site

# Prepare data for paired tests]
meta_paired <- meta %>%
  filter(site %in% c("Preen", "Cloaca")) %>%
  arrange(bird, site) %>%
  group_by(bird) %>%
  filter(n() == 2) %>%
  ungroup()

# Extract paired meta
observed_pre <- meta_paired %>% filter(site == "Preen") %>% pull(Observed)
observed_clo <- meta_paired %>% filter(site == "Cloaca") %>% pull(Observed)

shannon_pre <- meta_paired %>% filter(site == "Preen") %>% pull(Shannon)
shannon_clo <- meta_paired %>% filter(site == "Cloaca") %>% pull(Shannon)

simpson_pre <- meta_paired %>% filter(site == "Preen") %>% pull(Simpson)
simpson_clo <- meta_paired %>% filter(site == "Cloaca") %>% pull(Simpson)

# Perform the Paired t-Test for Observed species richness
t_test_result <- t.test(observed_pre, observed_clo, paired = TRUE)
print(t_test_result)

# Perform the Wilcoxon Signed-Rank Test for Shannon diversity
wilcox_test_shannon <- wilcox.test(shannon_pre, shannon_clo, paired = TRUE)
print(wilcox_test_shannon)

# Perform the Wilcoxon Signed-Rank Test for Simpson diversity
wilcox_test_simpson <- wilcox.test(simpson_pre, simpson_clo, paired = TRUE)
print(wilcox_test_simpson)
```


##Remove taxa with relative abundance <0.005%
```{r Remove Low Abundance Taxa}
minTotRelAbun = .00005
x = taxa_sums(physeq)
keepTaxa = (x / sum(x)) > minTotRelAbun
physeqprune = prune_taxa(keepTaxa, physeq)
physeqprune
```

##save physeq object as a file
```{r Save PhyloseqPrune Object}
##save physeq object as R file
save(physeqprune, file="../RData/physeqprune.RData")
```

##load physeq
```{r Load PhyloseqPrune Object}
load("../RData/physeqprune.RData")
```

##Number of shared ASVs site (found in 50% or more) 
```{r Identify Number of Shared ASVs}
sitevenn=ps_venn(
  physeqprune,
  "site",
  fraction = .50,
  weight = FALSE,
  relative = TRUE,
  plot = TRUE
)
sitevenn
```

##List of shared ASVs species (found in 50% or more) t=0
```{r List Shared ASVs}
sitelist=ps_venn(
  physeqprune,
  "site",
  fraction = .5,
  weight = FALSE,
  relative = TRUE,
  plot = FALSE
)
sitelist

# Extract the taxonomy table
tax_table <- as.data.frame(tax_table(physeqprune))

# Function to get genus and species for a list of ASVs
get_genus_species_for_asvs <- function(asv_list, tax_table) {
  # Subset the taxonomy table for the given ASVs
  matched_taxa <- tax_table[rownames(tax_table) %in% asv_list, ]
  
  # Function to find the most specific identified taxonomic level
  get_first_identified <- function(row) {
    if (!is.na(row["Species"]) && row["Species"] != "" && row["Species"] != "unidentified" && !is.na(row["Genus"]) && row["Genus"] != "" && row["Genus"] != "unidentified") {
      return(paste(row["Genus"], row["Species"], sep = " "))
    } else if (!is.na(row["Genus"]) && row["Genus"] != "" && row["Genus"] != "unidentified") {
      return(paste("Genus:", row["Genus"]))
    } else {
      tax_levels <- c("Family", "Order", "Class", "Phylum", "Kingdom")
      for (col in tax_levels) {
        if (!is.na(row[col]) && row[col] != "" && row[col] != "unidentified") {
          return(paste(col, row[col], sep = ": "))
        }
      }
    }
    return("unidentified")
  }
  
  # Apply the function to each row
  matched_taxa$First_Identified_Taxa <- apply(matched_taxa, 1, get_first_identified)
  
  # Return the ASV and First_Identified_Taxa columns
  return(data.frame(ASV = rownames(matched_taxa), First_Identified_Taxa = matched_taxa$First_Identified_Taxa))
}

# Get genus and species for each group in sitelist
genus_species_cloaca <- get_genus_species_for_asvs(sitelist$Cloaca, tax_table)
genus_species_preen <- get_genus_species_for_asvs(sitelist$Preen, tax_table)
genus_species_cloaca_preen <- get_genus_species_for_asvs(sitelist$Cloaca__Preen, tax_table)

# Save results to CSV files
write.csv(genus_species_cloaca, "../venntaxa/taxa_cloaca.csv", row.names = FALSE)
write.csv(genus_species_preen, "../venntaxa/taxa_preen.csv", row.names = FALSE)
write.csv(genus_species_cloaca_preen, "../venntaxa/taxa_cloaca_preen.csv", row.names = FALSE)
```

##export tiff with 300dpi
```{r Export Venn Diagram}
ggsave(
  filename="../figures/Figure02Venn.tiff",
  plot = sitevenn,
  width = 200,
  height = 200,
  units = c("mm"),
  dpi = 300,
  bg = "white"
)
```

##Bray Curtis Calculation
```{r Bray-Curtis Analysis, results='hide', message=FALSE, warning=FALSE}
set.seed(777)
dist = phyloseq::distance(physeqprune, method="bray", weighted=TRUE) #calculate Bray-Curtis dissimilarity
ordination = ordinate(physeqprune, method="NMDS", distance=dist) #perform ordination on distance matrix
```

##Bray Curtis Site Plot
```{r Plot ordination}
braysite=plot_ordination(physeq, ordination, color="site") + 
  theme_classic() +
  theme(strip.background = element_blank()) + stat_ellipse(aes(group=site)) + ggtitle("Bray-Curtis Dissimilarity") + theme(plot.title = element_text(size = 10, face = "bold", hjust = .5), axis.title.y = element_text(size = 10, face="bold"),
         axis.title.x = element_text(size = 10, face="bold"),) + 
  labs(color="Site") +
  scale_color_manual(values = c("Cloaca" = "#AA4499", "Preen" = "#88CCEE"))

braysite
```

##Bray Curtis Site Stats
```{r Statistics on Bray-Curtis}
adonis2(dist ~ sample_data(physeqprune)$site)
```

##Bray Curtis Species ANOSIM
```{r}
anosim <- data.frame(sample_data(physeqprune))
anosim(dist, anosim$site, permutations=9999)
```

##export tiff with 300dpi
```{r Export NMDS}
ggsave(
  filename="../figures/Figure03BetaDiv.tiff",
  plot = braysite,
  width = 250,
  height = 150,
  units = c("mm"),
  dpi = 300,
)
```

##Bar plots of Abundance per individual samples in site (Phylum-Merge <5%)
```{r Graph of Phylum Abundance}
physeq2 = filter_taxa(physeqprune, function(x) mean(x) > 0.05, TRUE)
physeq3 = transform_sample_counts(physeq2, function(x) x / sum(x) )
glom<-psmelt(physeq3)
glom <- tax_glom(physeq3, taxrank = 'Phylum')
data<-psmelt(glom)
data$Phylum <- as.character(data$Phylum) 
data$Phylum[data$Abundance < 0.05] <- "< 5% abund."
medians <- ddply(data, ~Phylum, function(x) c(median=median(x$Abundance)))
remainder <- medians[medians$median <= 0.05,]$Phylum
data[data$Phylum %in% remainder,]$Phylum <- "< 5% abund."
data$Phylum[data$Abundance < 0.05] <- "< 5% abund."
spatial_plot <- ggplot(data=data, aes(x=bird, y=Abundance, fill=Phylum)) + 
  facet_wrap(~site, scales = "free")
barplotphylum<-spatial_plot + geom_bar(aes(), stat="identity", position="fill") + 
  ggtitle("Phylum Abundance at Each Site") + 
  theme (axis.text.x = element_text(angle=90), 
         plot.title = element_text(size = 10, face = "bold", hjust = .5)) 
barplotphylum
```

##export tiff with 300dpi
```{r}
ggsave(
  filename="../figures/Figure00Barplot.tiff",
  plot = barplotphylum,
  width = 250,
  height = 150,
  units = c("mm"),
  dpi = 300,
)
```

##Bar plots of Abundance per individual samples in site (Family-Merge <4%)
```{r}
physeq2 = filter_taxa(physeqprune, function(x) mean(x) > 0.04, TRUE)
physeq3 = transform_sample_counts(physeq2, function(x) x / sum(x) )
glom<-psmelt(physeq3)
glom <- tax_glom(physeq3, taxrank = 'Family')
data<-psmelt(glom)
data$Family <- as.character(data$Family) 
data$Family[data$Abundance < 0.04] <- "< 4% abund."
medians <- ddply(data, ~Family, function(x) c(median=median(x$Abundance)))
remainder <- medians[medians$median <= 0.04,]$Family
data[data$Family %in% remainder,]$Family <- "< 4% abund."
data$Family[data$Abundance < 0.04] <- "< 4% abund."
spatial_plot <- ggplot(data=data, aes(x=bird, y=Abundance, fill=Family)) + 
  facet_wrap(~site, scales = "free")
barplotfamily<-spatial_plot + geom_bar(aes(), stat="identity", position="fill") + 
  ggtitle("Family Abundance at Each Site") + 
  xlab("Bird") +
  theme (axis.text.x = element_text(angle=90),
         axis.title.y = element_text(face="bold"),
         axis.title.x = element_text(face="bold"),
         plot.title = element_text(size = 10, face = "bold", hjust = .5)) 
barplotfamily
```
##Bar plots of Abundance per individual samples in site (Order-Merge <5%)
```{r}
physeq2 = filter_taxa(physeqprune, function(x) mean(x) > 0.05, TRUE)
physeq3 = transform_sample_counts(physeq2, function(x) x / sum(x) )
glom<-psmelt(physeq3)
glom <- tax_glom(physeq3, taxrank = 'Order')
data<-psmelt(glom)
data$Order <- as.character(data$Order) 
data$Order[data$Abundance < 0.05] <- "< 5% abund."
medians <- ddply(data, ~Order, function(x) c(median=median(x$Abundance)))
remainder <- medians[medians$median <= 0.05,]$Order
data[data$Order %in% remainder,]$Order <- "< 5% abund."
data$Order[data$Abundance < 0.05] <- "< 5% abund."
spatial_plot <- ggplot(data=data, aes(x=bird, y=Abundance, fill=Order)) + 
  facet_wrap(~site, scales = "free")
barplotorder<-spatial_plot + geom_bar(aes(), stat="identity", position="fill") + scale_fill_manual(values = c("#888888", "#CC6677", "#88CCEE", "#117733", "#332288", 
                               "#AA4499", "#44AA99", "#999933", "#882255", "#661100", 
                               "#6699CC", "#DDCC77")) +
  ggtitle("Order Abundance at Each Site") + 
  xlab("Bird") +
  theme (axis.text.x = element_text(angle=90),
         axis.title.y = element_text(face="bold"),
         axis.title.x = element_text(face="bold"),
         plot.title = element_text(size = 10, face = "bold", hjust = .5)) 
barplotorder
```

##export tiff with 300dpi
```{r}
ggsave(
  filename="../figures/Figure00BarplotOrder.tiff",
  plot = barplotorder,
  width = 250,
  height = 150,
  units = c("mm"),
  dpi = 300,
)
```

#Differential species ID
```{r warning=FALSE}
# Extract abundance data (OTU table)
abundance_data <- as.data.frame(otu_table(physeqprune))

# Extract the grouping variable (site) from sample data
site_group <- sample_data(physeqprune)$site

# Ensure the grouping variable is a factor
site_group <- as.factor(site_group)

# Run SIMPER analysis
simper_result <- simper(abundance_data, group = site_group, permutations = 100)

# Extract SIMPER results for each pairwise comparison
simper_summary <- summary(simper_result)

# Extract the data frame from the list
simper_df <- simper_summary$Preen_Cloaca

# Convert the data frame to include species names
simper_df <- as.data.frame(simper_df)
simper_df$species <- rownames(simper_df)
rownames(simper_df) <- NULL

# Add comparison information
simper_df$comparison <- "Preen_Cloaca"

# Save SIMPER results to a CSV file
write.csv(simper_df, "../differentialexpression/simper_results.csv", row.names = FALSE)

# Extract OTU names and their contributions
otu_contributions <- simper_df %>%
  select(species, average, sd, ratio, ava, avb, cumsum, comparison) %>%
  arrange(species)

# Perform Mann-Whitney U Test on OTU contributions between groups
results <- data.frame()

for (otu in unique(otu_contributions$species)) {
  # Subset the data for the current OTU
  otu_data <- subset(otu_contributions, species == otu)
  
  # Perform Mann-Whitney U Test
  test_result <- wilcox.test(otu_data$ava, otu_data$avb)
  
  # Store results
  results <- rbind(results, data.frame(OTU = otu, p.value = test_result$p.value))
}

# Adjust p-values for multiple testing (optional)
results$adj.p.value <- p.adjust(results$p.value, method = "BH")

# Save Mann-Whitney results to a CSV file
write.csv(results, "../differentialexpression/mann_whitney_results.csv", row.names = FALSE)

# View significant results
significant_results <- subset(results, adj.p.value < 0.05)

# Save significant results to a CSV file
write.csv(significant_results, "../differentialexpression/significant_results.csv", row.names = FALSE)

# Print significant results
print(significant_results)
```

##grid alpha div
```{r}
grid=ggarrange(barplotorder,                                                 
          ggarrange(barsite, braysite, ncol = 2, labels = c("B", "C")), 
          nrow = 2, 
          labels = "A"                                        
          ) 

grid
```

##export tiff with 300dpi
```{r}
ggsave(
  filename="../figures/FigureGrid.tiff",
  plot = grid,
  width = 250,
  height = 200,
  units = c("mm"),
  dpi = 300,
)
```
